\bf Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-03 paper.qmd
Starting to add robust statistics application

>>>>

Another application would compare outlying points to a reference normal distribution that is derived using robust methods on the original data. To illustrate this we work on the weather data example from @filzmoser2018. The data contains 16 numerical measurements that are averages across the three summer months June, July and August, these are provided for 68 years (1955 - 2022), see @filzmoser2018 for more details.

To estimate the underlying normal distribution the data is first centered and scaled using the median and the median absolute deviation (MAD), before applying the minimum covariance determinant (MCD) estimator @rousseeuw1985multivariate using the implementation in @robustbase. The MCD estimates for the mean vector and the variance-covariance matrix are then used to define the reference normal distribution.

Here we will consider points to be outlying if they are more than $5 \sigma$ away from that mean value, given $p=16$ this corresponds to a Mahalanobis distance of $60$ or larger. This will identify $20$ out of the $68$ observations as outlying. Since these are outlying in different combinations of variables, as found in @filzmoser2018, we will further separate the outlying points into clusters. To focus on the direction when grouping the points we first normalize observations to have length $1$ and then apply k-means clustering. Considering different cluster validation statistics computed from @fpc suggests that $k=4$ or $k=5$ is preferred, for simplicity we will work with $k=4$.

We first apply the new anomaly index to the full dataset, such that the final projection will identify a compromise and provide a global picture of where the outlying points differ from the normal distribution.

```{r}
#| eval: false
#| echo: false
library(ShapleyOutlier)
library(robustHD)
library(tidyverse)
library(tourr)
library(fpc)

data("WeatherVienna")
# first part of the code is preprocessing
# this is following the vignette from the package
# see https://cran.r-project.org/web/packages/ShapleyOutlier/vignettes/ShapleyOutlier_examples.html
weather_summer <- WeatherVienna %>% dplyr::select(-c(`t`, t_max, t_min, p_max, p_min)) %>%
  drop_na() %>%
  filter(month %in% c("JUN", "JUL", "AUG")) %>%
  filter(year >= 1955) %>%
  group_by(year) %>%
  dplyr::select(-month) %>%
  summarise(across(.cols = everything(), function(x) mean(x)))

X <- weather_summer %>% dplyr::select(-c(num_frost, num_ice, year))
rownames(X) <- weather_summer$year
#> Warning: Setting row names on a tibble is deprecated.
X <- robStandardize(X)

set.seed(1)
MCD <- covMcd(X, alpha = 0.5, nsamp = "best")
#> Warning in .fastmcd(x, h, nsamp, nmini, kmini, trace = as.integer(trace)): 'nsamp = "best"' allows maximally 100000 subsets;
#> computing these subsets of size 17 out of 68
mu <-MCD$center
Sigma <- MCD$cov

###### preprocessing done
# this is now our own code to work with these inputs

# start with finding outlying points and clustering
size_weather <- qchisq(2*pnorm(5)-1, ncol(X))
dist_pd <- mahalanobis(X, mu, Sigma)
X_outside <- X[dist_pd > size_weather,]
X_outside_scaled <- t(apply(X_outside, 1,
                    function(x) (x)/sqrt(sum(x^2))))
for(k in 2:6){
  set.seed(150)
  X_km <- kmeans(X_outside_scaled, k)
  cs <- cluster.stats(dist(X_outside_scaled), X_km$cluster)
  print(cs$dunn)
}

#preferred solution 5 clusters using dunn/dunn2 index
set.seed(150)
X_km <- kmeans(X_outside_scaled, 5)
# making a vector that labels all points
# use 0 to indicate point is not an outlier
cluster_labels_all <- as_tibble(X,
                               rownames = "year") %>%
  left_join(as_tibble(X_km$cluster,
                              rownames = "year")) %>%
  mutate(cluster = if_else(is.na(value), 0, value))
cluster_labels <- factor(cluster_labels_all$cluster,
                         levels = c(1:5, 0))

set.seed(129)
animate_xy(X, 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           axes="off", col = as.factor(cluster_labels),
           palette = "Plasma", half_range = 5,
           center = FALSE, rescale = FALSE)

```

### 2024-07-03 paper.tex
Starting to add robust statistics application

>>>>

Another application would compare outlying points to a reference normal
distribution that is derived using robust methods on the original data.
To illustrate this we work on the weather data example from
\citet{filzmoser2018}. The data contains 16 numerical measurements that
are averages across the three summer months June, July and August, these
are provided for 68 years (1955 - 2022), see \citet{filzmoser2018} for
more details.

To estimate the underlying normal distribution the data is first
centered and scaled using the median and the median absolute deviation
(MAD), before applying the minimum covariance determinant (MCD)
estimator \citet{rousseeuw1985multivariate} using the implementation in
\citet{robustbase}. The MCD estimates for the mean vector and the
variance-covariance matrix are then used to define the reference normal
distribution.

Here we will consider points to be outlying if they are more than
\(5 \sigma\) away from that mean value, given \(p=16\) this corresponds
to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
out of the \(68\) observations as outlying. Since these are outlying in
different combinations of variables, as found in \citet{filzmoser2018},
we will further separate the outlying points into clusters. To focus on
the direction when grouping the points we first normalize observations
to have length \(1\) and then apply k-means clustering. Considering
different cluster validation statistics computed from \citet{fpc}
suggests that \(k=4\) or \(k=5\) is preferred, for simplicity we will
work with \(k=4\).

We first apply the new anomaly index to the full dataset, such that the
final projection will identify a compromise and provide a global picture
of where the outlying points differ from the normal distribution.
